\chapter{Conclusions and Future Work}
\label{chap:conclusiones}

\section{Summary of Goals Achievements}

This thesis introduced \textbf{CF-RALA}, an efficient extension of the ChangeFormer architecture for multi-temporal building change detection (BCD) in optical satellite imagery. The central objective was to explore whether a reduced-rank linear attention mechanism could replace standard quadratic self-attention without significantly degrading model performance. The proposed approach was implemented through the integration of Rank-Augmented Linear Attention (RALA) inside the Siamese encoder of ChangeFormer, resulting in a model that maintains global contextual reasoning while reducing computational complexity.

The experimental evaluation on the LEVIR-CD and DSIFN-CD benchmarks showed that CF-RALA preserves the predictive capacity of the baseline in structured urban environments. On LEVIR-CD, the model reproduced the F1, IoU, and OA metrics of ChangeFormer-256 with differences that remain within the typical variation observed in state-of-the-art methods. On DSIFN-CD, which contains heterogeneous spectral conditions and spatial irregularities, CF-RALA achieved competitive recall and F1 values, while presenting a moderate decrease in IoU due to sensitivity to fine-scale boundaries. Despite this, CF-RALA was able to localize the most relevant structural changes, validating the hypothesis that a low-rank linear approximation of attention can serve as a viable alternative for efficient BCD.

From a computational perspective, the model demonstrated improved scalability with respect to both resolution and batch size. CF-RALA allowed training on larger batch configurations and higher-resolution patches under the memory constraints of a 40\,GB NVIDIA A100 GPU, an outcome not feasible with the original quadratic attention. This enhancement is particularly relevant for very-high-resolution imagery such as PeruSAT-1, for which efficient Transformer-based models are essential. Overall, the results confirm that the integration of low-rank linear attention provides a favorable balance between efficiency and accuracy, opening the door to scalable urban monitoring systems.

\section{Limitations}

Although the findings of this thesis are promising, several limitations emerged throughout the research process. A major constraint was the inherent memory cost of Transformer architectures. Even with RALA, full-resolution experiments at $1024 \times 1024$ pixels remained computationally demanding, limiting the exploration of end-to-end models directly on complete satellite scenes. Patch-based evaluation, while standard in literature, restricts the assessment of contextual consistency across large spatial extents.

Dataset variability also introduced significant challenges. LEVIR-CD provided clean, well-aligned urban scenes, allowing clear validation of architectural changes. However, DSIFN-CD introduced illumination differences, vegetation dynamics, cloud interference, and subtle temporal misalignments that complicated model evaluation. These sources of semantic noise are not always reflected in ground-truth labels, making metrics such as IoU and OA more difficult to interpret and potentially masking real model capabilities or weaknesses.

Another important limitation was the absence of a Peruvian multitemporal building-change dataset using PeruSAT-1 or similar sensors. Due to this constraint, model evaluation relied exclusively on international benchmarks that differ from the radiometric and morphological characteristics of Peruvian urban areas. Consequently, the applicability of CF-RALA to the national context could not be empirically validated. Lastly, the integration of RALA into the ChangeFormer architecture required extensive adjustments to ensure tensor compatibility, stable training dynamics, and accurate computation of FLOPs and parameters. These difficulties highlight the complexity of adapting efficient attention mechanisms to hierarchical Transformer models.

\section{Future Work}

Future research inspired by this thesis can advance along three complementary directions: improving model design, developing local datasets, and moving toward operational deployment with PeruSAT-1 imagery.

From a methodological standpoint, a natural extension involves exploring hybrid architectures where certain encoder levels retain full self-attention while others employ RALA or similar linear mechanisms. Such selective allocation of computational resources may preserve expressive power at critical scales while maintaining overall efficiency. Further advances in low-rank approximations, kernel-based attentions, or dynamic rank selection could yield even better performance. Incorporating self-supervised or contrastive pretraining on large corpora of unlabeled urban imagery also represents a promising path, potentially strengthening the model's feature representations before fine-tuning.

With regard to data, the creation of a multitemporal building-change benchmark for Peru is a critical next step. Developing a dataset based on PeruSAT-1 or other VHR sensors, with carefully annotated building changes, would enable evaluation directly aligned with the intended application of this research. Establishing labeling protocols, spatial sampling strategies, and suitable evaluation metrics would provide a valuable contribution to the national and regional remote sensing community.

Finally, from an operational perspective, integrating CF-RALA into a complete urban monitoring workflow offers a highly impactful line of work. Such a system would encompass geometric and radiometric preprocessing of PeruSAT-1 imagery, change detection inference at scale, mosaicking, and the generation of cartographic products for public institutions. Research into deployment-oriented strategies—such as model compression, distributed inference, or cloud-native implementations—would further enhance practicality. Complementing satellite imagery with auxiliary data, such as digital elevation models or cadastral layers, may also help distinguish true structural changes from surface or illumination variations.

Taken together, these future directions seek to strengthen the methodological, data-driven, and operational foundations needed to build efficient, deployable remote sensing tools for large-scale urban analysis in the Peruvian context.
