\section{\uppercase{Method}}
\label{sec:method}

RALA-ChangeFormer is constructed by replacing every MHSA block in the \textbf{encoder} of ChangeFormer with the rank-augmented linear attention module described in Section~\ref{subsec:rala}, where attention-related computational costs are dominant. All remaining components-CNN encoder, feed-forward layers, skip connections, and multi-scale fusion-remain unchanged, preserving full architectural compatibility with the baseline design. The integration mechanism is summarized in Figure~\ref{fig:rala_block}.

Given the token sequence $X^{(l)} \in \mathbb{R}^{N_l \times C_l}$ at resolution level $l$, the block first computes query, key, and value projections
\begin{equation}
Q = X^{(l)}W_Q,\qquad
K = X^{(l)}W_K,\qquad
V = X^{(l)}W_V,
\label{eq:qkv_projection}
\end{equation}

after which the attention operation proceeds exactly following the equations introduced in Section~\ref{subsec:rala}, namely: the global query descriptor (Eq.~\ref{eq:global_query}), the token-wise relevance coefficients (Eq.~\ref{eq:alpha_softmax}), the rank-augmented key–value buffer (Eq.~\ref{eq:rala_rank_aug}), and the modulated output features (Eq.~\ref{eq:Yi_modulated}).

The resulting token-specific attention output $A$ replaces the MHSA output of the original ChangeFormer block and is incorporated via a residual connection:
\begin{equation}
X_{\text{out}}
    = X^{(l)} + \mathrm{LN}(A),
\qquad
Y = \mathrm{FFN}(X_{\text{out}}),
\label{eq:residual_ffn}
\end{equation}

This construction preserves the overall representational structure of the original ChangeFormer architecture-including multi-scale refinement, long-range reasoning, and residual learning-while substituting the quadratic-cost MHSA blocks \textbf{exclusively within the encoder} with a linear-time rank-augmented mechanism.

\noindent\textbf{Computational Complexity.}
Let $N$ denote the number of tokens and $d$ the hidden dimension.  
Standard MHSA requires \textbf{time: }$\mathcal{O}(N^{2}d)$ and \textbf{memory: }$\mathcal{O}(N^{2})$ due to the formation of the dense $N\times N$ attention matrix.  
In contrast, the RALA module scales as \textbf{time: }$\mathcal{O}(Nd)$ and \textbf{memory: }$\mathcal{O}(d^{2})$ because no pairwise token interactions are constructed explicitly.  
Since $N \gg d$ for VHR remote sensing images, this substitution results in substantially reduced computational cost and enables significantly larger batch sizes, faster training, and improved scalability to higher spatial resolutions.

Functionally, the architecture preserves all capabilities of ChangeFormer—multi-scale feature reasoning, long-range spatial interactions, and precise boundary refinement—while reducing the computational burden within each Transformer block. Because RALA operates as a drop-in replacement for MHSA, the model maintains full compatibility with pretraining strategies, multi-scale inference, and existing training pipelines. This makes RALA-ChangeFormer suitable for high-resolution BCD and scalable to large bi-temporal datasets under fixed GPU budgets.