\documentclass[a4paper,twoside]{article}

\usepackage{epsfig}
\usepackage{subcaption}
\usepackage{calc}
\usepackage{amssymb}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{multicol}
\usepackage{pslatex}
\usepackage{algorithm2e}
\usepackage[bottom]{footmisc}
\usepackage{SCITEPRESS}
\usepackage{cite}
\usepackage{hyperref}
\usepackage[none]{hyphenat}
\usepackage[table]{xcolor}
\definecolor{pastelorange}{HTML}{F5D08A}
\begin{document}

\title{Efficient Multi-Temporal Building Change Detection with Reduced-Rank Linear Attention}

\author{\authorname{Ricardo A. Acuña-Villogas\sup{1}\orcidAuthor{0009-0005-7543-2232}, Germain García-Zanabria\sup{4}\orcidAuthor{0000-0003-3266-9043}, Cristian Lopez\sup{3}\orcidAuthor{0000-0002-2568-650X} and Rensso Mora-Colque\sup{2}\orcidAuthor{0000-0003-4734-8752}}
\affiliation{\sup{1}\sup{2}\sup{3}\sup{4}Department of Data Science, University of Engineering and Technology, Barranco, Lima, Peru}
\email{\{ricardo.acuna,  ggarciaz, clopezd, rmora\}@utec.edu.pe}
}

\keywords{Building Change Detection, Linear Attention, Remote Sensing, Computational Efficiency, Satellite Imagery}

\abstract{
Transformer-based architectures have become the dominant paradigm for bi-temporal building change detection (BCD). However, their application to very high–resolution (VHR) imagery remains limited by the quadratic complexity of multi-head self-attention (MHSA). This work introduces RALA-ChangeFormer, an efficient variant of ChangeFormer that replaces the encoder’s MHSA modules with rank-augmented linear attention (RALA). The integration preserves the hierarchical structure of the original model and maintains its ability to perform bi-temporal spatio–temporal reasoning. Experiments on the LEVIR-CD and DSIFN-CD benchmarks show that RALA-ChangeFormer delivers segmentation accuracy comparable to the baseline, with variations within standard training variance. The method reduces attention-related GFLOPs, lowers peak GPU memory consumption, and enables substantially larger batch sizes at multiple resolutions. Ablation studies confirm that the efficiency gains arise primarily from the encoder, where token counts are highest. Overall, the results demonstrate that rank-augmented linear attention provides a practical accuracy–efficiency trade-off for VHR BCD. The proposed approach improves computational scalability without sacrificing predictive performance, making high-resolution or large-scale training more accessible under fixed hardware constraints.
}

\onecolumn \maketitle \normalsize \setcounter{footnote}{0} \vfill

\input{1_Introduction}
\input{2_Related_Work}
\input{3_Background}
\input{4_Method}
\input{5_Experiments}
\input{6_Conclusions}


\section*{\uppercase{Acknowledgements}}

We gratefully acknowledge the support of the GINIA research group, where this work was initiated. GINIA provided an academic research environment that supported the early stages of this work in urban data analysis and covered the conference registration fee.

%\textit{$\backslash$section*\{ACKNOWLEDGEMENTS\}}

\bibliographystyle{IEEEtran}
{\small
\bibliography{example}}

\end{document}